---
title: "Untitled"
author: "Claire Meuter"
date: "2023-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(jsonlite)
library(here)
library(sf)
library(terra)
library(rnaturalearth)
```

## Accessing IUCN API

- first: add API key to .Renviron and why do it this way?
    - in console: `usethis::edit_r_environ()`
    - add line with a useful recognizable name and your API key
    - e.g., `IUCN_KEY=12345678`
- second: accessing API key
    - now in your code, you can access this key with `api_key <- Sys.getenv('IUCN_KEY')`
- if no personal API key yet, use the API demo one: 
    - 9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee

```{r}
api_key <- "9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee"
```


#access APi dataset version
```{r}
domain_name <- 'http://apiv3.iucnredlist.org' #putting in the domain to access the data 
version_end <- "api/v3/version"

version_url <- file.path(domain_name, version_end) #stitching these together into a web address

api_version <- jsonlite::fromJSON(version_url)

```
## how many species are in the ICUN dataset? 

```{r}
count_stem <- 'api/v3/speciescount?token=%s' #%s is where I can substitute a string

count_end <- sprintf(count_stem, api_key)

count_url <- file.path(domain_name, count_end)

spp_count <- jsonlite::fromJSON(count_url)

as.integer(spp_count$count) #Gives us the count of the species 
```

##get a page of results 
```{r}
page_stem <- "api/v3/species/page/%s?token=%s"
page_end <- sprintf(page_stem, 1, api_key)

page_url <- file.path(domain_name, page_end)

spp_page1 <- fromJSON(page_url)

spp_df <- spp_page1$result %>% 
  mutate(category = factor(category, levels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX'))) %>% 
  filter(!is.na(category))

ggplot(spp_df) +
  geom_bar(aes(x = category, fill = category)) +
  scale_fill_brewer(palette = 'RdYlGn', direction = -1) + theme_minimal()
```


## get current and historical assesment information for particular species 
```{r}
hist_stem <- 'api/v3/species/history/name/%s?token=%s'
spp <- "Dermochelys%20coriacea" #leatherback turtle %20 tells web this is a space. Same as putting a space in 

hist_end <- sprintf(hist_stem, spp, api_key)

hist_url <- file.path(domain_name, hist_end)

spp_hist <- fromJSON(hist_url)

spp_hist_df <- spp_hist$result

### this is as far as we got in lab-- more can be done later on my own. 
```


### Map threatened species 
```{r}
#loading in species info 
iucn_spp_info <- read_csv(here('data/iucn_marine_spp_info_2021-3.csv'))

iucn_spp_info$cat %>% unique() #to see the types of categories

cell_ids <- read_csv(here('data/am_latlong.csv')) 
spp_cells <- read_csv(here('data/am_spp_cells.csv'))
spp_ids <- read_csv(here('data/am_spp_ids.csv')) #match up species id to name 
#we want to combine these ci]olumns in a meaningful way so that we can map them to lat and long 

spp_risk_cells <- spp_cells %>% 
  inner_join(cell_ids, by = 'loiczid') %>% 
  inner_join(spp_ids, by = "am_sid") %>% 
  inner_join(iucn_spp_info, by=c('binomial' = 'sciname'))

threaten_pct_cells <- spp_risk_cells %>% 
  filter(prob >= 0.5) %>% 
  mutate(threatened = (tolower(cat) %in% c('vu','en','cr'))) %>% 
  filter(cat != 'dd') %>% 
  group_by(lon, lat) %>% 
  summarize(pct_threatened = sum(threatened)/n())
#now we can turn this into a raster 

```
#making this data a raster 
```{r}
spp_risk_rast <- terra::rast(threaten_pct_cells, type = "xyz", crs = "epsg:4326")

plot(spp_risk_rast)


```
## now we'll plot with ggplot 
```{r}
#showing how to easily make a df from raster
spp_rast_df <- spp_risk_rast %>% 
  as.data.frame(xy = TRUE)


p <- ggplot(threaten_pct_cells) +
  geom_raster(aes(x = lon, y = lat, fill = pct_threatened)) +
  scale_fill_viridis_c() +
  coord_sf()
p

```
## make it fancy 
```{r}
land_sf <- rnaturalearth::ne_countries(scale = 50, 
                                       country = c("united states of america", 'mexico'),
                                       returnclass = "sf")

land_sf_cropped <- land_sf %>% 
  st_crop(spp_risk_rast)

p2 <- p + 
  geom_sf(data = land_sf_cropped,
          fill = 'grey80', color = 'red', alpha = .5) + theme_void()

p2
```

